The Talking Heads Experiment

### 2015-04-10 chap12/chap12.tex
quotation marks for chapters 8 to 10

The previous chapters have shown that huge advances have been made in the decade after the original Talking Heads experiment. 
It is now realistic to do experiments with physical humanoid robots, something which was unthinkable 
in the late nineties. Whereas categorization using 
basic perceptually grounded categories was a big achievement at the end of the nineties, experiments now routinely use 
complex compositional semantics. Another big boost has come from the development of a grammar formalism
(Fluid Construction Grammar) that is adequate from the viewpoint of linguistic complexity and allows at the same time
sophisticated experiments in grammar emergence. We now have several in-depth studies of language strategies for several 
domains although much more work obviously needs to be done to map out the vast landscape of language strategies found in 
human languages. This work is difficult and time consuming because each semantic domain generates its own set of fundamental 
problems in perception and action, conceptual representation, grammar, and language game interaction.

@@

To clarify these issues, let us first define a distinction between emergence, language change, and evolution:\is{culture-driven language
evolution}
\begin{itemize}
\item {\itshape Language emergence} is the phenomenon whereby a language system, i.e. a set of conventions and conceptualizations
for a particular semantic domain, for example tense-aspect, color, space or determination, arises in a population and 
becomes shared, based on a particular strategy (or a set of strategies) that is shared by all agents. 
\item {\itshape Language change} happens when the components of a language system and its conceptualizations shift, but the 
change remains within the bounds of a particular strategy. For example, the boundaries how different cases are used 
may shift (as is happening with the dative and genitive at the moment in German) or a new basic color category and
color word may arise and claim a region within the color space (as happened with the word ``orange" which 
started to be used as a basic color word in 16th century Middle English to occupy a distinctive region between yellow and red). 
\item {\itshape Language evolution} occurs when new strategies arise. There is plenty of evidence from creole languages and 
from historical linguistics that this happens at a cultural time scale. Here are three examples:
\begin{enumerate}
\item Phrase structure (in the sense of 
strict word order based on abstract hierarchical patterns) emerged gradually in 
proto-Indo European, from a stage with largely free word order but a strong inflectional system \citep{VandeVelde:2011}. 
\item Word order and prepositions became the dominant strategy instead of morphological cases 
for expressing argument structure in the late Old English period \citep{VanKemenade:1987}. 
\item There is currently an evolution going on in Spanish clitics (``le", ``la", ``lo") whereby the etymological system 
of Standard Spanish, which uses clitics to express different cases (nominative, dative, accusative), is shifting 
to a referential system (expressing gender and number). Case differentiation is lost, but existing forms are recruited
for the new functions implied by a referential strategy  \citep{Fernandez-Ordonez:1999}. 
\end{enumerate}
\end{itemize}
It is not always obvious from the observation of language facts whether we are dealing with change (within the boundaries 
of a strategy) or with a new strategy. However, this becomes entirely clear when we try to computationally model the processes of 
acquisition, expansion or invention of the language strategy involved. We always try to understand change first using existing 
strategies until we are forced to introduce a new strategy because the differences at the conceptual or linguistic 
level are too profound to computationally handle them.

@@

Today, Darwinian evolutionary dynamics \is{Darwinian evolutionary dynamics}
is recognized as a general principle that can and has been applied to 
many different types of systems, in chemistry, neurobiology or 
economics.\footnote{See \cite{Arthur:1996}, \cite{Edelman:1987} and \cite{Luisi:2001}.}
It has also been used successfully to come up 
with new artificial systems, for example to evolve controllers for robots using 
genetic algorithms \citep{Nolfi:2000}. It is therefore not so surprising that 
selectionism can also be mapped to the cultural and more specifically to the linguistic domain in the following way: 
\begin{enumerate}
\item Speakers generate variants, either accidentally through errors in their speech production, or deliberately, 
for example, when they try to express a novel idea by inventing new forms or 
appropriating existing forms for new purposes, for example, by coercing a word or grammatical contruction into a new usage. 
\item Which variants will survive in the communal language depends on selectionist criteria. I will 
discuss them shortly in more detail. They include for example whether the speaker is understood by the hearer, or whether 
it reduces the cognitive complexity of parsing and producing and hence allows larger sentences to be understood. 
\item The self-enforcing causal loop happens when variants created by one speaker start to propagate in the population 
and begin to replace their competitors. For example, a particular word to express a given meaning gets out of fashion 
and is taken over by another one. Some of these developments are simply fashion trends, but the more profound ones, 
in particular the ones that involve grammar, are related to critical selectionist criteria. 
\end{enumerate}
This Darwinian dynamic is already active in the operation of language strategies, because a strategy regulates the 
creation and selection of variants, determines how feedback has to be handled, and implies a self-enforcing causal loop through 
lateral inhibition so that variants satisfying the selection criteria become 
dominant in the population.

@@

Linguists are often reticent to use a selectionist approach because they believe that language is 
maladapted \citep{Boeckx:2005}. However we have to keep in mind that: (i) An evolutionary process 
seldom comes up with the perfect solution, even though it is remarkable effective given that there is no 
enlightened intelligent designer. (ii) There are conflicting criteria so that there is no optimal solution. 
For example, the speaker may want to minimize the effort needed for producing and pronouncing a sentence but this 
may make it harder for the hearer to reconstruct an interpretation of the meaning. (iii) Some of the selectionist 
criteria have to do with processing. A linguistic theory that only focuses on competence has nothing to say about 
processing and therefore cannot lay bare how processing issues play a role in language evolution. (iv) It is often possible 
to maintain communicative success with a simpler system (even a purely lexical language without any grammar at all) and then 
gradually optimize semantic interpretation by the progressive introduction of more grammar - as we see indeed 
in the emergence of creole languages which regenerate syntactic complexity from 
a lexical language \citep{Mufwene:2001}.

What kind of selectionist forces are relevant in the case of language? \is{selectionist forces}
Clearly, the bottom line is that language users want to achieve {\bfshape communicative success}. 
Communicative success means that the speaker achieves his or her non-linguistic goal. For example, if a speaker
wants a hearer to perform some action (such as ``Can you get me the car keys?"), there is communicative success if 
the hearer indeed performs the requested action. Speakers and hearers generally do not care whether their sentences
are perfectly grammatical or complete or semantically accurate. Speakers assume that hearers are intelligent enough
to infer the missing information and that they can fill in the required details from the context. 
In normal spoken dialogue, communicative success is immediately obvious, either 
from the actions that the hearer performs as a consequence of an utterance or by gestural cues whereby 
the hearer informs the speaker that he or she is being understood.

@@

{\bfshape Cognitive effort}: In order to cope with the incredible speed of normal language production and comprehension, 
reducing cognitive effort is of primordial importance. Cognitive effort is expended at all levels
of language: How much time and memory 
needs to be spent in coming up with adequate conceptualizations? How complex is the process of constructing
a sentence? How 
difficult is it to articulate the speech sounds? How efficiently can the sentence be parsed? How complicated is the 
interpretation of the sentence? If sentences are too 
complex to comprehend, for example because they trigger a large amount of combinatorial search, then 
hearers give up. Or if sentences 
are too hard to produce, the production process will be too slow to maintain the hearer's attention. 
Optimizations of cognitive effort are often in conflict for speaker and hearer. What requires less effort for the 
speaker might mean more effort for the hearer and vice versa. 
\is{cognitive effort}

{\bfshape Learnability}: Speakers must occasionally expand their language systems to express novel meanings or 
better capture the attention of hearers with a `fresh' way of saying something. But these innovations will 
lead nowhere if they cannot be acquired by hearers. They will lead to communicative failure and 
the innovation will not propagate in the population and therefore it has no chance to survive in the shared language.
Learnability depends on whether speaker and hearer share the same language strategies, how much additional context 
is available, and how far the speaker has stretched established conventions to achieve his communicative goal. 
\is{learnability}

{\bfshape Social conformity}: Speakers and hearers can greatly optimize their language production and 
comprehension by making their language systems as similar as possible. Moreover, details of language use, such 
as pronunciation, signal to which group a speaker belongs, and speakers often seek social conformity with the 
language use of their peers in order to belong to the group. Or they mimic the language systems 
of the most prestigious group in order to show they belong to that. The opposite happens as well. Speakers 
may try to mark themselves as linguistically different from a group to which they do not want to belong. 
\is{social conformity}

@@

To illustrate this approach let us look at a first example, \is{German article system}
using a fascinating study conducted by Remi van Trijp of change in the German article system as historically 
observed over the past 10 centuries.\footnote{
This section draws on the following publications: \cite{VanTrijp:2013} and \cite{VanTrijp:2013b}.}
The change is puzzling, because the article system started from a rather clean 
design in old High German to retain only a few forms in contemporary German. This contemporary system 
seems to be maladapted because the same forms are now used for many different purposes (syncretism) causing 
uncertainty and hence combinatorial search in parsing. For example, in Old High German there was a three-way gender 
distinction between masculine, neuter, and feminine for both nominative and accusative plurals, but in New 
High German only one form ``die" is left. 
Many linguists have argued that these changes are just random, caused by phonological erosion and confusion 
of closely related word-forms. But van Trijp hypothesized that the changes 
were actually optimizations, whereby speakers optimized their own cognitive effort at the cost of that of
hearers, even though the reduction in system complexity stopped 
at a critical point to preserve the disambiguation power of the system. See \figref{fig:german}.

@@

Van Trijp initialized three populations of agents: with the Old High German system, 
the Middle High German system, or the New High German System. He then defined the following four dimensions for the 
fitness landscape: 
\begin{enumerate}
\item {\itshape Search effort} which is the amount of search that is needed, scaled between minimum (when there is one marker for every of 
the 18 possible configurations and so no search is needed) and maximum (when there are no markers at all).
\item {\itshape Disambiguation power} which is based on the number of possible interpretations left after application of the grammar. 
The disambiguation power is also scaled with maximum power for 12 markers and minimum for 0 markers (not counting the genitive). 
\item {\itshape Articulatory effort} which is calculated in terms of the number of movements that articulators such as the lips and 
tongue have to make when pronouncing the different phonemes that make up each article. 
\item {\itshape Acoustic precision} which is based on the phonological distance between words.
\end{enumerate}
Criteria (1), (2) and (4) pertain to the hearer and (1) and (3) to the speaker. They are calculated for each utterance
in a corpus and the weighted average gives a value for the performance of the language system for each of these dimensions. The global 
performance can be assessed by taking a (weighted) average of each of the values for all dimensions. 
As expected, Old High German scores the highest for disambiguation power and acoustic precision and the lowest for articulatory effort 
and processing effort, thus being an advantage for the hearer but not for the speaker (see \figref{fig:german-system}).

On the other hand, the New High German system manages to become equally effective for disambiguation while requiring less search effort
and articulatory effort, if other complementary cues are made available to the hearer by additional grammatical 
strategies, namely grammatical agreement between the article and the noun within a noun phrase, so 
that features of the article can be deduced from the noun, and syntactic valence of the verb, because that suggests
which cases noun-phrases and hence their articles may have. 
This demonstrates that we should not always focus on a single strategy alone 
but have to take other relevant strategies into account. 
Moreover some forms, which are not very distinctive from an acoustic point of 
view (such as ``der", ``den" and ``dem"), and therefore risk being collapsed into a single form, are 
nevertheless maintained because they are needed to resolve ambiguities that are not mitigated by 
other syntactic devices. This shows how an evolutionary point of view helps us to formulate explanatory hypotheses about language.

@@

This example shows how it is possible to analyse the fitness landscape for stages of a language system at different points in 
time. The same analysis can also be done for different regional variants that are currently competing to become dominant 
in the standard language. An example of the latter is the currently ongoing evolution
from the etymological Spanish pronoun system to the referential pronoun systems, 
which has three different possible regional variants: ``le\'ismo", ``la\'ismo" and  ``lo\'ismo". 
Van Trijp showed how each of these variants could equally arise with the same referential strategy and how it is 
a matter of historical contingency which one becomes dominant. Indeed we see that different regions of Spain
have chosen different variants \citep{VanTrijp:2010}.

Here is another example of a fitness landscape from the domain of space, developed by 
Michael Spranger.\footnote{These results are documented at length in: \cite{Spranger:2014}.}
It focuses on understanding the role of grammar for optimizing spatial language. \is{Spatial Language Game}
The experiment involves QRIO humanoids
(see \figref{fig:spatial}) and uses a 
sophisticated vision system developed by Spranger, together with IRL and Fluid 
Construction Grammar (in the previous chapter). A variety of spatial language strategies 
were operationalised, in particular:

\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{chap11/figs/spatial}}
\caption{ Set-up for the spatial language game experiments. Two robots are looking at a scene 
with a few blocks, a bigger box with tags indicating front, back or left and right side, and a global landmark in the form 
of a large tag pasted on the wall. {\itshape Top left and right:} World as seen through the camera of the robot. 
{\itshape Bottom left and right:} The internal world model as perceived by the left and right robot.}
\label{fig:spatial}
\end{figure}
\begin{enumerate}
\item The use of {\itshape projective relations} such as front/back, above/below, left/right. This requires imposing a coordinate
system on the world model and name areas within this coordinate system. A common coordinate system is to take 
an object, such as the human body of the speaker as reference point and project vectors emanating from the body. 
\item The use of {\itshape proximal relations} such as far/near. This requires that the distances of objects with respect to the 
reference point are computed and categorized. 
\item The use of {\itshape absolute relations}, such as North/West/East/South, related to global directions. 
\item The use of an {\itshape allocentric landmark} (as opposed to an egocentric one)
from which the situation is conceptualised. For example, in the 
phrase ``the block in front of the box", the box is acting as a landmark and the projective spatial relations are 
computed as emanating from that object. Another example is ``the block on your left" where the landmark is the 
hearer. 
\end{enumerate}

Spranger analysed the fitness landscape for grammatical strategies. He compared 
one population of agents that was initialized with a pidgin German, i.e. only a vocabulary for objects and spatial relations 
but no grammar, and another population initialized with a full locative grammar of German. He investigated three external 
factors: 
\begin{enumerate}
\item Scenes with many objects distributed around a landmark.
\item Scenes with several objects but no landmark that can be used as a spatial vantage point. 
\item Scenes with an allocentric and without an allocentric landmark but more complex with respect to number 
of objects and distribution of objects compared to (1). 
\end{enumerate}

\begin{figure}
\centerline{\includegraphics[width=0.5\linewidth]{chap12/figs/why-grammar-german}}
\caption{ Investigation how the two language systems (lexical or grammatical) are performing with respect to the 
communicative success dimension and for each of the three ecological conditions. A purely lexical system in itself has 
already considerable success, but grammar becomes more and more relevant when the situation becomes more complex.
The bottom images show examples for the three ecological conditions being examined.}
\label{fig:why-grammar-german}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.5\linewidth]{chap12/figs/why-grammar-german-avg-interpretations}
\includegraphics[width=0.5\linewidth]{chap12/figs/why-grammar-conceptualization-interpretation-equality}}
\caption{ Left: performance for semantic ambiguity for the same three ecological conditions. We see that without grammar
the world model needs to be consulted heavily to weed out potential interpretations. And Right: Performance for interpretation sharing.
Only with grammar are agents able to share the same semantic structure, ensuring a higher degree of communicative success.}
\label{fig:ambiguity}
\end{figure}

@@

We now move one level up, and show how a fitness landscape can be identified at the level of strategies as well. Rather than supplying 
agents with language systems, we initialize them 
with specific strategies, let them build language systems, and then examine the performance 
of these language systems for the different dimensions of the fitness landscape, so as to understand which niche 
the strategy occupies. Here is one example of this approach 
from the domain of color (developed by Joris Bleys).\footnote{These results are documented at length in: \cite{Bleys:2014}.}

The previous chapter already showed that there are 
different strategies for color language: basic color terms, graded membership (``very green"), compounds (``sky blue"), 
brightness qualification (``light green"), etc. All these strategies use a particular color space, more specifically 
the CIE L*u*v* space where the L* dimension represents brightness (ranging from black to white), the u* 
dimension represents the red-green opponent channel, 
and the v* dimension the yellow-blue channel. This space was chosen 
because the distance between two colours in the L*u*v* space accurately represents 
the psychological distances between these colours as perceived by human subjects. \is{Color Description Game}

Let us focus on a family of strategies that all use a prototype-based approach, i.e. they perform categorization using 
a standard nearest neighbour classification algorithm operating over an inventory of prototypical points, 
as also used in the Proper Naming Game and the Basic Colour Naming game discussed in the previous chapter. 
Three variant strategies are imaginable (and can be found in human languages): 
\begin{enumerate}
\item The {\itshape brightness-only strategy} which uses only the L* dimension (dark, grey, bright, etc.).
\item The {\itshape hue-only strategy} which uses only the hue dimension (blue, yellow, green, etc.). 
\item And the {\itshape full-color strategy} which uses all dimensions. 
\end{enumerate}
In order to understand the niche of each strategy, Bleys has set up experiments where the context is controlled 
with respect to the importance of hue. 
For example, agents can be presented with contexts with samples where the hue is the same but brightness is different on the one 
hand versus contexts with samples where the brightness is the same but hue is different. In between we get a mixed 
importance of hue.

@@

Bleys initialized the three populations with the three different strategies and let them first learn an interpretation of 
the Spanish focal colors (with prototypes in the brightness-only, hue-only, and full-colour spaces). Then he measured 
performance for communicative success for the varying ecological conditions. 
Not unexpectedly, he found that depending on the percentage of the importance of hue differences in the context, a different strategy 
is more optimal (see \figref{fig:bar}). Not surprisingly, a hue-only strategy fails completely when the samples in each  
context have equal hue but different brightness, whereas a brightness-only strategy fails when the samples have 
equal brightness but different hue.

@@

Recall that three internal factors were used 
for the fitness landscape in this domain: communicative success, semantic ambiguity and interpretation sharing. Spranger started from 
a population without any grammar but just a lexicon for expressing spatial relations and properties of objects. 
The agents were then initialized with a strategy that would build a grammar consisting of hierarchical
constructions which group all the components of a spatial relation, put them in a particular sequential order,
and add a marker to them. For example, a construction could arise that groups a spatial relation (like `front'), 
a determiner acting as selector (like `the'), and a noun introducing an object class (e.g. ``box"), puts them 
in the order: spatial relation - determiner - noun, and adds a marker to them (e.g. ``-bo").

>>>>

The previous chapters have shown that huge advances have been made in the decade after the original Talking Heads experiment. 
It is now realistic to do experiments with physical humanoid robots, something which was unthinkable 
in the late nineties. Whereas categorisation using 
basic perceptually grounded categories was a big achievement at the end of the nineties, experiments now routinely use 
complex compositional semantics. Another big boost has come from the development of a grammar formalism
(Fluid Construction Grammar) that is adequate from the viewpoint of linguistic complexity and allows at the same time
sophisticated experiments in grammar emergence. We now have several in-depth studies of language strategies for several 
domains although much more work obviously needs to be done to map out the vast landscape of language strategies found in 
human languages. This work is difficult and time consuming because each semantic domain generates its own set of fundamental 
problems in perception and action, conceptual representation, grammar, and language game interaction.

@@

To clarify these issues, let us first define a distinction between emergence, language change, and evolution:\is{culture-driven language
evolution}
\begin{itemize}
\item {\itshape Language emergence} is the phenomenon whereby a language system, i.e. a set of conventions and conceptualisations
for a particular semantic domain, for example tense-aspect, color, space or determination, arises in a population and 
becomes shared, based on a particular strategy (or a set of strategies) that is shared by all agents. 
\item {\itshape Language change} happens when the components of a language system and its conceptualisations shift, but the 
change remains within the bounds of a particular strategy. For example, the boundaries how different cases are used 
may shift (as is happening with the dative and genitive at the moment in German) or a new basic color category and
color word may arise and claim a region within the color space (as happened with the word `orange' which 
started to be used as a basic color word in 16th century Middle English to occupy a distinctive region between yellow and red). 
\item {\itshape Language evolution} occurs when new strategies arise. There is plenty of evidence from creole languages and 
from historical linguistics that this happens at a cultural time scale. Here are three examples:
\begin{enumerate}
\item Phrase structure (in the sense of 
strict word order based on abstract hierarchical patterns) emerged gradually in 
proto-Indo European, from a stage with largely free word order but a strong inflectional system \citep{VandeVelde:2011}. 
\item Word order and prepositions became the dominant strategy instead of morphological cases 
for expressing argument structure in the late Old English period \citep{VanKemenade:1987}. 
\item There is currently an evolution going on in Spanish clitics (`le', `la', `lo') whereby the etymological system 
of Standard Spanish, which uses clitics to express different cases (nominative, dative, accusative), is shifting 
to a referential system (expressing gender and number). Case differentiation is lost, but existing forms are recruited
for the new functions implied by a referential strategy  \citep{Fernandez-Ordonez:1999}. 
\end{enumerate}
\end{itemize}
It is not always obvious from the observation of language facts whether we are dealing with change (within the boundaries 
of a strategy) or with a new strategy. However, this becomes entirely clear when we try to computationally model the processes of 
acquisition, expansion or invention of the language strategy involved. We always try to understand change first using existing 
strategies until we are forced to introduce a new strategy because the differences at the conceptual or linguistic 
level are too profound to computationally handle them.

@@

Today, Darwinian evolutionary dynamics \is{Darwinian evolutionary dynamics}
is recognised as a general principle that can and has been applied to 
many different types of systems, in chemistry, neurobiology or 
economics.\footnote{See \cite{Arthur:1996}, \cite{Edelman:1987} and \cite{Luisi:2001}.}
It has also been used successfully to come up 
with new artificial systems, for example to evolve controllers for robots using 
genetic algorithms \citep{Nolfi:2000}. It is therefore not so surprising that 
selectionism can also be mapped to the cultural and more specifically to the linguistic domain in the following way: 
\begin{enumerate}
\item Speakers generate variants, either accidentally through errors in their speech production, or deliberately, 
for example, when they try to express a novel idea by inventing new forms or 
appropriating existing forms for new purposes, for example, by coercing a word or grammatical contruction into a new usage. 
\item Which variants will survive in the communal language depends on selectionist criteria. I will 
discuss them shortly in more detail. They include for example whether the speaker is understood by the hearer, or whether 
it reduces the cognitive complexity of parsing and producing and hence allows larger sentences to be understood. 
\item The self-enforcing causal loop happens when variants created by one speaker start to propagate in the population 
and begin to replace their competitors. For example, a particular word to express a given meaning gets out of fashion 
and is taken over by another one. Some of these developments are simply fashion trends, but the more profound ones, 
in particular the ones that involve grammar, are related to critical selectionist criteria. 
\end{enumerate}
This Darwinian dynamic is already active in the operation of language strategies, because a strategy regulates the 
creation and selection of variants, determines how feedback has to be handled, and implies a self-enforcing causal loop through 
lateral inhibition so that variants satisfying the selection criteria become 
dominant in the population.

@@

Linguists are often reticent to use a selectionist approach because they believe that language is 
maladapted \citep{Boeckx:2005}. However we have to keep in mind that: (i) An evolutionary process 
seldom comes up with the perfect solution, even though it is remarkable effective given that there is no 
enlightened intelligent designer. (ii) There are conflicting criteria so that there is no optimal solution. 
For example, the speaker may want to minimise the effort needed for producing and pronouncing a sentence but this 
may make it harder for the hearer to reconstruct an interpretation of the meaning. (iii) Some of the selectionist 
criteria have to do with processing. A linguistic theory that only focuses on competence has nothing to say about 
processing and therefore cannot lay bare how processing issues play a role in language evolution. (iv) It is often possible 
to maintain communicative success with a simpler system (even a purely lexical language without any grammar at all) and then 
gradually optimise semantic interpretation by the progressive introduction of more grammar - as we see indeed 
in the emergence of creole languages which regenerate syntactic complexity from 
a lexical language \citep{Mufwene:2001}.

What kind of selectionist forces are relevant in the case of language? \is{selectionist forces}
Clearly, the bottom line is that language users want to achieve {\bfshape communicative success}. 
Communicative success means that the speaker achieves his or her non-linguistic goal. For example, if a speaker
wants a hearer to perform some action (such as `Can you get me the car keys?'), there is communicative success if 
the hearer indeed performs the requested action. Speakers and hearers generally do not care whether their sentences
are perfectly grammatical or complete or semantically accurate. Speakers assume that hearers are intelligent enough
to infer the missing information and that they can fill in the required details from the context. 
In normal spoken dialogue, communicative success is immediately obvious, either 
from the actions that the hearer performs as a consequence of an utterance or by gestural cues whereby 
the hearer informs the speaker that he or she is being understood.

@@

{\bfshape Cognitive effort}: In order to cope with the incredible speed of normal language production and comprehension, 
reducing cognitive effort is of primordial importance. Cognitive effort is expended at all levels
of language: How much time and memory 
needs to be spent in coming up with adequate conceptualisations? How complex is the process of constructing
a sentence? How 
difficult is it to articulate the speech sounds? How efficiently can the sentence be parsed? How complicated is the 
interpretation of the sentence? If sentences are too 
complex to comprehend, for example because they trigger a large amount of combinatorial search, then 
hearers give up. Or if sentences 
are too hard to produce, the production process will be too slow to maintain the hearer's attention. 
Optimisations of cognitive effort are often in conflict for speaker and hearer. What requires less effort for the 
speaker might mean more effort for the hearer and vice versa. 
\is{cognitive effort}

{\bfshape Learnability}: Speakers must occasionally expand their language systems to express novel meanings or 
better capture the attention of hearers with a ``fresh'' way of saying something. But these innovations will 
lead nowhere if they cannot be acquired by hearers. They will lead to communicative failure and 
the innovation will not propagate in the population and therefore it has no chance to survive in the shared language.
Learnability depends on whether speaker and hearer share the same language strategies, how much additional context 
is available, and how far the speaker has stretched established conventions to achieve his communicative goal. 
\is{learnability}

{\bfshape Social conformity}: Speakers and hearers can greatly optimise their language production and 
comprehension by making their language systems as similar as possible. Moreover, details of language use, such 
as pronunciation, signal to which group a speaker belongs, and speakers often seek social conformity with the 
language use of their peers in order to belong to the group. Or they mimic the language systems 
of the most prestigious group in order to show they belong to that. The opposite happens as well. Speakers 
may try to mark themselves as linguistically different from a group to which they do not want to belong. 
\is{social conformity}

@@

To illustrate this approach let us look at a first example, \is{German article system}
using a fascinating study conducted by Remi van Trijp of change in the German article system as historically 
observed over the past 10 centuries.\footnote{
This section draws on the following publications: \cite{VanTrijp:2013} and \cite{VanTrijp:2013b}.}
The change is puzzling, because the article system started from a rather clean 
design in old High German to retain only a few forms in contemporary German. This contemporary system 
seems to be maladapted because the same forms are now used for many different purposes (syncretism) causing 
uncertainty and hence combinatorial search in parsing. For example, in Old High German there was a three-way gender 
distinction between masculine, neuter, and feminine for both nominative and accusative plurals, but in New 
High German only one form `die' is left. 
Many linguists have argued that these changes are just random, caused by phonological erosion and confusion 
of closely related word-forms. But van Trijp hypothesised that the changes 
were actually Optimisations, whereby speakers optimised their own cognitive effort at the cost of that of
hearers, even though the reduction in system complexity stopped 
at a critical point to preserve the disambiguation power of the system. See \figref{fig:german}.

@@

Van Trijp initialised three populations of agents: with the Old High German system, 
the Middle High German system, or the New High German System. He then defined the following four dimensions for the 
fitness landscape: 
\begin{enumerate}
\item {\itshape Search effort} which is the amount of search that is needed, scaled between minimum (when there is one marker for every of 
the 18 possible configurations and so no search is needed) and maximum (when there are no markers at all).
\item {\itshape Disambiguation power} which is based on the number of possible interpretations left after application of the grammar. 
The disambiguation power is also scaled with maximum power for 12 markers and minimum for 0 markers (not counting the genitive). 
\item {\itshape Articulatory effort} which is calculated in terms of the number of movements that articulators such as the lips and 
tongue have to make when pronouncing the different phonemes that make up each article. 
\item {\itshape Acoustic precision} which is based on the phonological distance between words.
\end{enumerate}
Criteria (1), (2) and (4) pertain to the hearer and (1) and (3) to the speaker. They are calculated for each utterance
in a corpus and the weighted average gives a value for the performance of the language system for each of these dimensions. The global 
performance can be assessed by taking a (weighted) average of each of the values for all dimensions. 
As expected, Old High German scores the highest for disambiguation power and acoustic precision and the lowest for articulatory effort 
and processing effort, thus being an advantage for the hearer but not for the speaker (see \figref{fig:german-system}).

On the other hand, the New High German system manages to become equally effective for disambiguation while requiring less search effort
and articulatory effort, if other complementary cues are made available to the hearer by additional grammatical 
strategies, namely grammatical agreement between the article and the noun within a noun phrase, so 
that features of the article can be deduced from the noun, and syntactic valence of the verb, because that suggests
which cases noun-phrases and hence their articles may have. 
This demonstrates that we should not always focus on a single strategy alone 
but have to take other relevant strategies into account. 
Moreover some forms, which are not very distinctive from an acoustic point of 
view (such as `der', `den' and `dem'), and therefore risk being collapsed into a single form, are 
nevertheless maintained because they are needed to resolve ambiguities that are not mitigated by 
other syntactic devices. This shows how an evolutionary point of view helps us to formulate explanatory hypotheses about language.

@@

This example shows how it is possible to analyse the fitness landscape for stages of a language system at different points in 
time. The same analysis can also be done for different regional variants that are currently competing to become dominant 
in the standard language. An example of the latter is the currently ongoing evolution
from the etymological Spanish pronoun system to the referential pronoun systems, 
which has three different possible regional variants: `le\'ismo', `la\'ismo' and `lo\'ismo'. 
Van Trijp showed how each of these variants could equally arise with the same referential strategy and how it is 
a matter of historical contingency which one becomes dominant. Indeed we see that different regions of Spain
have chosen different variants \citep{VanTrijp:2010}.

Here is another example of a fitness landscape from the domain of space, developed by 
Michael Spranger.\footnote{These results are documented at length in: \cite{Spranger:2014}.}
It focuses on understanding the role of grammar for optimising spatial language. \is{Spatial Language Game}
The experiment involves QRIO humanoids
(see \figref{fig:spatial}) and uses a 
sophisticated vision system developed by Spranger, together with IRL and Fluid 
Construction Grammar (in the previous chapter). A variety of spatial language strategies 
were operationalised, in particular:

\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{chap11/figs/spatial.pdf}}
\caption{ Set-up for the spatial language game experiments. Two robots are looking at a scene 
with a few blocks, a bigger box with tags indicating front, back or left and right side, and a global landmark in the form 
of a large tag pasted on the wall. {\itshape Top left and right:} World as seen through the camera of the robot. 
{\itshape Bottom left and right:} The internal world model as perceived by the left and right robot.}
\label{fig:spatial}
\end{figure}
\begin{enumerate}
\item The use of {\itshape projective relations} such as front/back, above/below, left/right. This requires imposing a coordinate
system on the world model and name areas within this coordinate system. A common coordinate system is to take 
an object, such as the human body of the speaker as reference point and project vectors emanating from the body. 
\item The use of {\itshape proximal relations} such as far/near. This requires that the distances of objects with respect to the 
reference point are computed and categorised. 
\item The use of {\itshape absolute relations}, such as North/West/East/South, related to global directions. 
\item The use of an {\itshape allocentric landmark} (as opposed to an egocentric one)
from which the situation is conceptualised. For example, in the 
phrase `the block in front of the box', the box is acting as a landmark and the projective spatial relations are 
computed as emanating from that object. Another example is `the block on your left' where the landmark is the 
hearer. 
\end{enumerate}

Spranger analysed the fitness landscape for grammatical strategies. He compared 
one population of agents that was initialised with a pidgin German, i.e. only a vocabulary for objects and spatial relations 
but no grammar, and another population initialised with a full locative grammar of German. He investigated three external 
factors: 
\begin{enumerate}
\item Scenes with many objects distributed around a landmark.
\item Scenes with several objects but no landmark that can be used as a spatial vantage point. 
\item Scenes with an allocentric and without an allocentric landmark but more complex with respect to number 
of objects and distribution of objects compared to (1). 
\end{enumerate}

\begin{figure}
\centerline{\includegraphics[width=0.5\linewidth]{chap12/figs/why-grammar-german.pdf}}
\caption{ Investigation how the two language systems (lexical or grammatical) are performing with respect to the 
communicative success dimension and for each of the three ecological conditions. A purely lexical system in itself has 
already considerable success, but grammar becomes more and more relevant when the situation becomes more complex.
The bottom images show examples for the three ecological conditions being examined.}
\label{fig:why-grammar-german}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.5\linewidth]{chap12/figs/why-grammar-german-avg-interpretations.pdf}
\includegraphics[width=0.5\linewidth]{chap12/figs/why-grammar-conceptualisation-interpretation-equality.pdf}}
\caption{ Left: performance for semantic ambiguity for the same three ecological conditions. We see that without grammar
the world model needs to be consulted heavily to weed out potential interpretations. And Right: Performance for interpretation sharing.
Only with grammar are agents able to share the same semantic structure, ensuring a higher degree of communicative success.}
\label{fig:ambiguity}
\end{figure}

@@

We now move one level up, and show how a fitness landscape can be identified at the level of strategies as well. Rather than supplying 
agents with language systems, we initialise them 
with specific strategies, let them build language systems, and then examine the performance 
of these language systems for the different dimensions of the fitness landscape, so as to understand which niche 
the strategy occupies. Here is one example of this approach 
from the domain of color (developed by Joris Bleys).\footnote{These results are documented at length in: \cite{Bleys:2014}.}

The previous chapter already showed that there are 
different strategies for color language: basic color terms, graded membership (`very green'), compounds (`sky blue'), 
brightness qualification (`light green'), etc. All these strategies use a particular color space, more specifically 
the CIE L*u*v* space where the L* dimension represents brightness (ranging from black to white), the u* 
dimension represents the red-green opponent channel, 
and the v* dimension the yellow-blue channel. This space was chosen 
because the distance between two colours in the L*u*v* space accurately represents 
the psychological distances between these colours as perceived by human subjects. \is{Color Description Game}

Let us focus on a family of strategies that all use a prototype-based approach, i.e. they perform categorisation using 
a standard nearest neighbour classification algorithm operating over an inventory of prototypical points, 
as also used in the Proper Naming Game and the Basic Colour Naming game discussed in the previous chapter. 
Three variant strategies are imaginable (and can be found in human languages): 
\begin{enumerate}
\item The {\itshape brightness-only strategy} which uses only the L* dimension (dark, grey, bright, etc.).
\item The {\itshape hue-only strategy} which uses only the hue dimension (blue, yellow, green, etc.). 
\item And the {\itshape full-color strategy} which uses all dimensions. 
\end{enumerate}
In order to understand the niche of each strategy, Bleys has set up experiments where the context is controlled 
with respect to the importance of hue. 
For example, agents can be presented with contexts with samples where the hue is the same but brightness is different on the one 
hand versus contexts with samples where the brightness is the same but hue is different. In between we get a mixed 
importance of hue.

@@

Bleys initialised the three populations with the three different strategies and let them first learn an interpretation of 
the Spanish focal colors (with prototypes in the brightness-only, hue-only, and full-colour spaces). Then he measured 
performance for communicative success for the varying ecological conditions. 
Not unexpectedly, he found that depending on the percentage of the importance of hue differences in the context, a different strategy 
is more optimal (see \figref{fig:bar}). Not surprisingly, a hue-only strategy fails completely when the samples in each  
context have equal hue but different brightness, whereas a brightness-only strategy fails when the samples have 
equal brightness but different hue.

@@

Recall that three internal factors were used 
for the fitness landscape in this domain: communicative success, semantic ambiguity and interpretation sharing. Spranger started from 
a population without any grammar but just a lexicon for expressing spatial relations and properties of objects. 
The agents were then initialised with a strategy that would build a grammar consisting of hierarchical
constructions which group all the components of a spatial relation, put them in a particular sequential order,
and add a marker to them. For example, a construction could arise that groups a spatial relation (like `front'), 
a determiner acting as selector (like `the'), and a noun introducing an object class (e.g. `box'), puts them 
in the order: spatial relation - determiner - noun, and adds a marker to them (e.g. `-bo').
